### 2013-06-09 doc/pgas13/evaluation.tex
incorporating Mark's feedback and stuff

Figure~\ref{fig:vector} shows the results of the throughput experiments for the global Stack and Queue. Results are shown with flat combining completely disabled, only combining at the master core (``centralize''), and combining locally (``distributed'').

>>>>

Figure~\ref{fig:vector} shows the results of the throughput experiments for the global Stack and Queue. Results are shown with flat combining completely disabled, only combining at the master core (``centralized''), and combining locally (``distributed'').

### 2013-06-09 doc/pgas13/flat-combining.tex
incorporating Mark's feedback and stuff

Reduced synchronization comes from delegating work to another core.
To illustrate the issue, consider the shared synchronous queue, shown in Figure~\ref{fig:fc_shared_mem}, with pre-allocated storage and a \texttt{top} pointer protected by a lock. Without flat combining, whenever a thread attempts to push something on the stack, it must acquire the lock, put its value into the storage array, bump the top, and then release the lock. When many threads contend for the lock, all but one will fail and will have to retry. Each try forces a memory fence for consistency and consumes bandwidth, and as the number of threads increases, the fraction of successes plummets. Under flat combining, instead, threads add requests to the publication list. They each try to acquire the lock, and the one that succeeds becomes the combiner. Instead of retrying, the rest spin waiting for their request to be filled. The combiner walks the publication list, performs all of the requests, and when done, releases the lock. This has greatly reduced the synchronization on the stack's lock, but has introduced a new point of synchronization---threads must synchronize to add to the publication list. However, if a thread performs multiple operations on the stack, it can leave its publication record in the list and amortize the synchronization cost.

The above example of delegation is compelling in itself. However, the crux of the prior work is that data structure-specific optimization can be done to perform the combined operations more efficiently than separately.
As the combiner walks the publication list, it merges each non-empty publication record into a combined operation. In the case of the stack example shown in Figure~\ref{fig:fc_shared_mem}, as it walks the list, Thread 4 keeps track of the operations on its own temporary stack. When it encounters Thread 2's pop, it recognizes that it can satisfy that pop immediately with the push it just processed from Thread 3, so it fills each of their records and allows each of them to proceed. After traversing the rest of the publication list, the thread applies the combined operation to the actual data structure, in this case, the remaining two elements are pushed onto the top of the stack, and the requesting threads' records are marked as complete.
In the case of the stack, combining came in the form of matched pushes and pops, but many data structures have other ways in which operations can be matched locally.

@@

To illustrate how flat combining can be applied to Grappa, we must first describe what a global data structure might look like. Figure~\ref{fig:stacknofc} shows a simple PGAS translation of the shared-memory stack from earlier. A storage array is allocated out of the global heap, so its elements are striped across all the cores in the system. One core is designated the ``master'' to enforce global synchronization, and holds the elements of the data structure that all concurrent accessors must agree on, in this case, the \texttt{top} pointer.

All tasks that want to access the global stack hold a \texttt{GlobalAddress} to the Master object. To perform an operation, the task invokes a custom delegate operation that, like the \texttt{read()} delegate described earlier, blocks the task until complete. Example code to do a \texttt{push} is shown in Figure~\ref{fig:push}, the task must send a message to the master to acquire the lock. If it succeeds, it follows the \texttt{top} pointer, writes its new value to the end of the stack, returns to bump the \texttt{top} pointer and release the lock, and finally sends a message back to wake the calling worker. All other workers performing operations block at the first message until the lock is released. Because of Grappa's user-level blocking mechanisms, blocked requests do not have to retry or busy-wait. However, all thousand workers on each core all perform this synchronization and serialize on the master core.

@@

\paragraph{Distributed Combining}
Instead of all workers on a core sending independent synchronization messages and putting the burden all on the master, what if each core did its own combining first, and synchronized with the master in bulk?
Distributing the combining to each core allows the majority of the work to be performed in parallel and without communication.
As with shared-memory combining, distributed combining improves locality, reduces the amount of global synchronization, and allows operations to be composed in data structure-specific ways for even greater performance.
% allows synchronization to happen in parallel on all the cores, 
% Flat combining in Grappa is about allowing synchronization and  distributing synchronization 
% In a way, Grappa's automatic message aggregation is already trying to reduce synchronization traffic; however, the generic mechanism for serializing and packing has a cost, and in the end, the operations must still serialize on the master.
% Flat combining is about ``teaching'' it how to more efficiently perform many operations remotely in certain special cases.

% In order for workers to cooperate locally, a local ``proxy'' object is allocated in the core-private heap. This proxy serves as the

In distributed flat-combining, each core needs its own publication list to track all of its outstanding operations to be performed on the global data structure. In Grappa, this takes the form of a local \emph{proxy} object allocated from the core-local heap.
Conceptually, workers add their requests to the local publication list, one is chosen to do the combining, and the rest block until their request is satisfied.
However, because workers are scheduled cooperatively, each worker has atomic access to the local proxy ``for free'', so an explicit publication list is unnecessary.
Instead, each worker performs its own combining, pushing the 
, and blocks if it was unable to satisfy its request immediately. Figure~\ref{fig:stackfc} shows how in the global stack example, pushes and pops can be matched locally, avoiding global synchronization for those operations.

After all local combining has been done, one requesting worker is chosen to perform the combined operation globally. In Figure~\ref{fig:stackfc} again, Worker 4 becomes the combiner and performs much the same synchronization as in the un-combined case, but is able to push multiple elements with one synchronization. As in the uncombined case, the global lock must still be acquired, and concurrent combined requests must still block and serialize on the master, though a technique later will describe how flat combining can be applied here, too. Combining in this way allows synchronization to be distributed and performed in parallel, coarsening the granularity of the actual global synchronization and reducing the amount of serialization that occurs.

% In most cases, this means communicating with whichever core holds the ``lock-protected'' field that maintains sequential consistency. In the case of a global stack, this is the ``master'' \emph{top} pointer which must be incremented on pushes and decremented on pops. On the core where global synchronization is occurring, there must be synchronization between the multiple synchronization requests arriving from other cores.

% At this point we have traded off one remote synchronization per operation for an additional local synchronization per operation and some number fewer remote synchronizations (depending on the amount of combining that occurs). This would likely result in a net benefit due to local synchronization being inherently cheaper. However, because of the way the Grappa runtime schedules computation, it is actually possible to get the correct mutual exclusion/atomicity without any additional synchronization overhead.

% By design, each core in Grappa operates independently of the other cores. Workers are scheduled cooperatively, so each worker can assume atomicity until it performs an operation that yields to the scheduler, such as a remote call. Therefore workers within a core can cooperate without any explicit synchronization operations requiring memory fences. This allows them to publish requests to the local proxy very inexpensively, eliminating the need for a complex synchronization-amortizing queue. This has the added benefit that there will be no unused publication records, which could have become a problem with thousands of workers. Further, remote operations are processed through the same cooperative-multithreading scheduler on the remote side, so combined operations are trivially serialized and atomic.

\paragraph{Memory Consistency Model}

In the context of this work, sequential consistency guarantees that within a
task, operations will be in task order and that all tasks will observe the
same global order of operations. An example is shown in
Figure~\ref{fig:sync}.\TODO{make sync example}

The Grappa memory model is essentially the same as the C++ memory
model~\cite{boehm:drf0,N2480,N2800} (guaranteeing sequential
consistency for data-race free programs). To be conservative, delegate
operations block the calling worker until they have completed globally,
ensuring they can be used for synchronization. As such, delegate operations
within a task are guaranteed to be globally visible in program order and all
tasks observe the same global order.
Operations on synchronized data structures must provide the same guarantees.
Because it is not immediately obvious
that this distributed version of flat combining preserves sequential
consistency, we will now argue why it does so.

% \begin{itemize}
%   \item Block the calling worker until the operation is guaranteed to have a valid global order.
%   \item Local combining must not allow workers to observe a different order than is committed globally.
% \end{itemize}
To preserve sequentially-consistent semantics for flat-combined operations, they must obey a consistent global order.
To start with, local combined operations must have a serializable order among each other; this is unchanged from shared-memory flat-combining, and is trivially true due to the atomicity enabled by cooperative multithreading.
When combined operations are applied globally, they are serialized at a particular core that owns the synchronization (the ``master'' core for the Queue and Stack, or the corresponding hash cell for the Set and Map). The global order, then, is essentially the concatenation of the cores' sequential orders.
For this ``concatenation'' to be valid, the order observed by workers during local combining must be the same as what can be observed globally as operations are committed.

In the case of the Stack and Queue, this guarantee comes from applying a batch of push or pop operations atomically in a contiguous block on the global data structure. The second level of combining at the master preserves this as well by only allowing updates to proceed when they will not conflict and blocking all of them until they have completed.
Matching pushes and pops locally is one exception to the rule that operations must block until globally committed. Because a pop ``destroys'' the corresponding push, these operations together are independent of all others and can conceptually be placed anywhere in a total global ordering, so they can be safely matched locally.

The guarantees of combining set and map operations is more nuanced.
Insert and lookup operations performed by different tasks are inherently unordered unless synchronized externally (i.e., we assume data-race freedom).
Therefore, a batch of these operations can be committed globally in parallel, since they are guaranteed not to conflict with each other.
Note that, if an insert finds that its key is already in the combiner, it does not need to send its own message. However, it must still block until that insert is done, otherwise it may assume that the key has been inserted and perform some other operation that requires the insert to have happened, effectively reordering operations.
This ensures that if one wanted to use the insert to signal synchronization, it is guaranteed to have been committed globally before the task continues.

@@

% (wrong!) A lookup can be satisfied locally provided that the insert it relies on is guaranteed to eventually be committed globally. This allows the task to continue. If it does another lookup to a different key that is satisfied locally, then that task has observed one particular ordering of the two inserts. When those inserts commit globally, they will proceed in parallel to their respective hashed locations. Another task could

% The combining operations that are performed locally among requests on a single core must preserve the illusion of sequential order; this is unchanged from shared-memory flat-combining. When combined operations are applied globally, they are serialized at the ``master'' core for a given synchronization context. The global order is essentially the concatenation of each core's batch of serialized accesses.
% In the case where operations are satisfied locally (by matching up with other local operations), by definition these must be independent of other operations, so they can conceptually be placed anywhere in the global ordering (or even ignored completely in the global ordering).

% In the spirit of the Data-Race-Free-0 Model (guaranteeing sequential consistency for data-race free programs), operations on the global data structure are considered to be made atomic by a conceptual global ``lock'' on the data structure.

@@

\section{Grappa FC Framework}
% One of the strengths of flat combining is that the same combining mechanism can be applied in multiple data structures. Each structure must only specify how its operations compose.
% Many different data structures can integrate flat combining into their operation using the same delegation mechanism, making it easier to implement concurrent data structures.
In order to leverage the flat-combining paradigm in Grappa, we implemented a generic framework which can be used to improve performance for a number of global data structures. The FC framework handles the common problems of managing the combiners, handling worker wake-ups, and maintaining progress. When hooking into the framework, each data structure implementation need only define how to combine operations and globally commit them.

% As discussed before, Grappa's distributed combining takes a different approach to expressing how operation locally  flat combining design takes a different approach to performing 
The Grappa FC framework takes a different approach than the original flat-combining work for expressing how operations combine. As mentioned before, the local ``publication list'' is trivial to access atomically in Grappa due to workers being cooperatively scheduled.
Instead, in the Grappa FC Framework, a \emph{proxy} object is instantiated ahead of time on each core and each worker merges its update into that structure before blocking or becoming the combiner.
% In the original work, an explicit publication list is built up, and when the combiner scans over the list, it has a structure that it uses to accumulate requests in a logical way. In the case of the stack, for instance, it is a thread-local stack, which it can use to immediately match up pushes and pops. In the Grappa implementation, each worker's actions are atomic anyway, so rather than creating an explicit request list, the combiner object (or \emph{proxy}) is instantiated ahead of time for the data structure, and each worker does its own local combining into that structure before blocking. This reduces the space blowout of a large dynamically-sized publication list (with associated caching issues), and does not overly affect the expression of combining.
In this style, each global data structure must define the following:
\begin{enumerate}
  \item A local \emph{proxy} data structure for tracking updates to the global structure.
  \item \emph{Combining methods} that operate on the local proxy.
  \item A \emph{sync} method that globally commits the proxy's state.
\end{enumerate}
An example proxy declaration for the GlobalStack is shown in Figure~\ref{fig:proxy}.

@@

void push(GlobalAddress<GlobalStack> stack, T e) {
  // perform operation on the stack's master core
  delegate::call(stack.core(), [stack,e]{
    // force other operations to wait
    stack->lock.acquire();
    // do remote write to the top of the stack
    // (so other workers may interleave)
    delegate::write(stack->top, e);
    // bump the top pointer
    stack->top++;
    stack->lock.release();
  }); // blocks until response arrives
}\end{lstlisting}
\caption{Snippet of code for pushing to a global stack (without flat combining).}
\label{fig:push}
\end{figure}

@@

While a combining operation is in flight, new requests are likely to continue to accumulate. The framework transparently wraps the proxy object so that when it starts a sync on one, it can direct requests to a fresh combiner. It is important to note that combiners must start out fresh each time---it would be unsound for them to ``remember'' the previous updates and attempt to combine locally with them.

% The flat combining framework transparently wraps the \emph{proxy} object. Behind this layer of indirection, it manages multiple copies of the proxy so that, as discussed above, there is always at least one being committed and one to receive additional combining operations.

% explain simple non-combined global data structures in Grappa section...

\subsection{Global Stack and Queue}
Figure~\ref{fig:proxy} shows an excerpt from the definition of the proxy object for the Grappa GlobalStack. The proxy tracks pushes locally in the \texttt{pushed\_values} array. When \texttt{pop()} is called, if there are pushes in the local array, it immediately takes the value and wakes the last worker to push. Otherwise, it adds a pointer to a location on its stack where the \texttt{sync} operation, will eventually write the result. Because of local matching, when a Stack proxy is synchronized, it will have either all pops or all pushes, which makes the implementation of sync straightforward. Note that the operation to synchronize a batch of pushes looks almost identical to the code to do a single push from Figure~\ref{fig:push}.

@@

\subsection{Global Set, and Map}
The Grappa GlobalSet uses a simple chaining design, implemented with a global array of cells (allocated from the global heap), which are partitioned evenly among all the cores, and indexed with a globally-agreed hash function. Our flat-combining version supports both \texttt{insert} and \texttt{lookup}. To track all of the keys waiting to be inserted, we use the hash set implementation from the C++11 standard library (\texttt{std::unordered\_set}). Similar to the case with pops, for lookups, we must keep track of pointers for where to put the results of lookup operations, which the proxy does with a hash map (again from the C++ standard library) from keys to a list of pointers. As discussed in Section~\ref{sec:sc}, matching lookups with inserts locally would force a particular sequential order, so instead we only allow matching inserts with inserts and lookups with lookups. This allows \texttt{sync} to simply issue all of the inserts and lookups in parallel and block until all have completed.

Our implementation of the GlobalMap matches that of the Set, but of course stores values of a given type. This does not change the requirements of the proxy except that the return values of lookups must be a copy of the value rather than just a boolean.

>>>>

Reduced synchronization comes from delegating work to another core.
To illustrate the issue, consider the shared synchronous stack shown in Figure~\ref{fig:fc_shared_mem}, with pre-allocated storage and a \texttt{top} pointer protected by a lock. Without flat combining, whenever a thread attempts to push something on the stack, it must acquire the lock, put its value into the storage array, bump the top, and then release the lock. When many threads contend for the lock, all but one will fail and will have to retry. Each attempt forces an expensive memory fence and consumes bandwidth, and as the number of threads increases, the fraction of successes plummets. Under flat combining, instead, threads add requests to the publication list. They each try to acquire the lock, and the one that succeeds becomes the combiner. Instead of retrying, the rest spin on their request waiting for it to be filled. The combiner walks the publication list, performs all of the requests, and when done, releases the lock. This has greatly reduced the synchronization on the stack's lock, but has introduced a new point of synchronization---threads must synchronize to add to the publication list. However, if a thread performs multiple operations on the stack, it can leave its publication record in the list and amortize the synchronization cost. This publication list mechanism can be re-used in different data structures, saving each one from needing to implement its own clever synchronization mechanism.

The above example of delegation is compelling in itself. However, the crux of the prior work is that data structure-specific optimization can be done to perform the combined operations more efficiently than separately.
As the combiner walks the publication list, it merges each non-empty publication record into a combined operation. In the case of the stack example shown in Figure~\ref{fig:fc_shared_mem}, as it walks the list, Thread 4 keeps track of the operations on its own temporary stack. When it encounters Thread 2's pop, it recognizes that it can satisfy that pop immediately with the push it just processed from Thread 3, so it fills both of their records and allows them to proceed. After traversing the rest of the publication list, the thread applies the combined operation to the actual data structure, in this case, the two unmatched pushes are added to the top of the stack, and the requesting threads' records are marked as complete.
In the case of the stack, combining came in the form of matched pushes and pops, but many data structures have other ways in which operations can be matched.

@@

To illustrate how flat combining can be applied to Grappa, we must first describe what a global data structure looks like. Figure~\ref{fig:stacknofc} shows a simple PGAS translation of the shared-memory stack from earlier. A storage array is allocated out of the global heap, so its elements are striped across all the cores in the system. One core is designated the ``master'' to enforce global synchronization, and holds the elements of the data structure that all concurrent accessors must agree on, in this case, the \texttt{top} pointer.

All tasks that want to access the global stack hold a \texttt{GlobalAddress} to the Master object. To perform an operation, the task invokes a custom delegate operation that, like the \texttt{read} delegate described earlier, blocks the task until complete. Example code to do a \texttt{push} is shown in Figure~\ref{fig:push}, the task must send a message to the master to acquire the lock. If it succeeds, it follows the \texttt{top} pointer, writes its new value to the end of the stack, returns to bump the \texttt{top} pointer and release the lock, and finally sends a message back to wake the calling worker. All other workers performing operations block at the first message until the lock is released. Because of Grappa's user-level blocking mechanisms, blocked requests do not have to retry or busy-wait. However, all workers on each core perform this synchronization and serialize on a single core, causing that core to become the bottleneck.

@@

\paragraph{Distributed Combining.}
Instead of all workers on a core sending independent synchronization messages and putting the burden all on one core, what if each core did its own combining first, and synchronized with the master in bulk?
Distributing the combining to each core allows the majority of the work to be performed in parallel and without communication.
As with shared-memory combining, distributed combining improves locality, reduces the amount of global synchronization, and allows operations to be composed in data structure-specific ways for even greater performance.

% In a way, Grappa's automatic message aggregation is already trying to reduce synchronization traffic; however, the generic mechanism for serializing and packing has a cost, and in the end, the operations must still serialize on the master.
% Flat combining is about ``teaching'' it how to more efficiently perform many operations remotely in certain special cases.

In distributed flat-combining, each core builds its own publication list to track all the operations waiting to be committed to the global data structure.
In Grappa, for each global data structure, a local \emph{proxy} object is allocated from the core-local heap to intercept requests.
Conceptually, workers add their requests to a local publication list in the proxy, then one is chosen to do the combining, and the rest block until their request is satisfied.
However, because Grappa's scheduler is non-preemptive, each worker has atomicity ``for free'' until it performs a blocking operation (such as communication).
This means that an explicit publication list with clever synchronization is unnecessary.
Instead, workers perform their own combining, merging their operations directly into the local proxy, and block if unable to satisfy their requests immediately.
The proxy structure is chosen to be able to compactly aggregate operations and efficiently perform matching in cases where it is allowed. Figure~\ref{fig:stackfc} shows how in the global stack example, pushes and pops are matched on the proxy's local stack.

After all local combining has been done, one requesting worker is chosen to commit the combined operation globally. Back to the example in Figure~\ref{fig:stackfc}, Worker 4 becomes the combiner and performs much the same synchronization as in the uncombined case, but is able to push multiple elements with one synchronization. The global lock must still be acquired, so concurrent combined requests (from different cores) must block and serialize on the master. Combining in this way allows synchronization to be distributed and performed in parallel, coarsening the granularity of the actual global synchronization and reducing the amount of serialization that occurs.

\paragraph{Memory Consistency Model}
\label{sec:memory-model}
In the context of this work, sequential consistency guarantees that within a
task, operations will be in task order and that all tasks will observe the
same global order of operations.

The Grappa memory model is essentially the same as the C++ memory
model~\cite{boehm:drf0,N2480,N2800} (guaranteeing sequential
consistency for data-race free programs). To be conservative, delegate
operations block the calling worker until they have completed globally,
ensuring they can be used for synchronization. As such, delegate operations
within a task are guaranteed to be globally visible in program order and all
tasks observe the same global order.
Operations on synchronized data structures must provide the same guarantees.
Because it is not immediately obvious
that this distributed version of flat combining preserves sequential
consistency, we will now argue why it does.

To preserve sequentially-consistent semantics for flat-combined operations, they must obey a consistent global order.
To start with, local combined operations must have a serializable order among each other; this is unchanged from shared-memory flat-combining, and is trivially true due to the atomicity enabled by cooperative multithreading.
When combined operations are applied globally, they are serialized at a particular core that owns the synchronization (the ``master'' core for the Stack).
% haven't discussed queue, set, or map yet...
%(the ``master'' core for the Queue and Stack, or the corresponding hash cell for the Set and Map). The global order, then, is essentially the concatenation of the cores' sequential orders.
For this ``concatenation'' to be valid, the order observed by workers during local combining must be the same as what can be observed globally as operations are committed.

In the case of a stack or queue, this guarantee comes from applying a batch of push or pop operations atomically in a contiguous block on the global data structure. The second level of combining at the master preserves this as well by only allowing updates to proceed when they will not conflict and blocking all of them until they have completed.
For the \emph{stack}, matching pushes and pops locally is one exception to the rule that operations must block until globally committed. Because a pop ``destroys'' the corresponding push, these operations together are independent of all others and can conceptually be placed anywhere in a total global ordering, so they can be safely matched locally.

Combining set and map operations expose more nuances in the consistency guarantees.
Insert and lookup operations performed by different tasks are inherently unordered unless synchronized externally.
Therefore, a batch of these operations can be committed globally in parallel, since they are guaranteed not to conflict with each other.
Note that, if an insert finds that its key is already in the combiner, it does not need to send its own message. However, it must still block until that insert is done, otherwise it may assume that the key has been inserted and perform some other operation that requires the insert to have happened, effectively reordering operations.
This ensures that if one wanted to use the insert to signal synchronization, it is guaranteed to have been committed globally before the task continues.

@@

Whenever a global commit starts, a fresh combiner must be created for subsequent operations to use. If they were to use the old combiner's state to satisfy requests locally, it would violate global ordering because the local object would not reflect other cores' updates. This guarantees, for example, that destructive operations only happen once (i.e., a worker cannot locally pop the same value that another core's combined operation does).

@@

\section{Grappa FC Framework}
In order to leverage the flat-combining paradigm in Grappa, we implemented a generic framework which can be used to improve performance for a number of global data structures. The FC framework handles the common problems of managing the combiners, handling worker wake-ups, and maintaining progress. When hooking into the framework, each data structure implementation need only define how to combine operations and globally commit them.

The Grappa FC framework takes a different approach than the original flat-combining work for expressing how operations combine. As mentioned before, the local ``publication list'' is trivial to access atomically in Grappa's cooperative scheduling.
Instead, in the Grappa FC Framework, a \emph{proxy} object is instantiated ahead of time on each core and each worker merges its update into that structure before blocking or becoming the combiner.
In this style, each global data structure must define the following:
\begin{enumerate}
  \item A local \emph{proxy} data structure for tracking updates to the global structure.
  \item \emph{Combining methods} that operate on the local proxy.
  \item A \emph{sync} method that globally commits the proxy's state.
\end{enumerate}
An example proxy declaration for the GlobalStack is shown in Figure~\ref{fig:proxy}.

@@

void push(GlobalAddress<GlobalStack> stack, T e) {
  // perform operation on the stack's master core
  delegate::call(stack.core(), [stack,e]{
    // force other operations to wait
    stack->lock.acquire();
    // do remote write to the top of the stack
    // (so other workers may interleave)
    delegate::write(stack->top, e);
    // bump the top pointer
    stack->top++;
    stack->lock.release();
  }); // blocks until response arrives
}
\end{lstlisting}
\caption{Snippet of code for pushing to a global stack (without flat combining).}
\label{fig:push}
\end{figure}

@@

While a combining operation is in flight, new requests are likely to continue to accumulate. The framework transparently wraps the proxy object so that when it starts a \texttt{sync} on one, it can direct requests to a fresh combiner.
This is done by hiding instances of proxy objects behind a C++ smart-pointer-like object which provides the pointer to the current combiner, and whenever it detects that it should send, allocates a new combiner and points subsequent references to that.
It is important to note that combiners must start out fresh each time---it would be unsound for them to ``remember'' the previous updates and attempt to combine locally with them. We refer readers to Section~\ref{sec:memory-model} for why this must be so.

\subsection{Global Stack and Queue}
Figure~\ref{fig:proxy} shows an excerpt from the definition of the proxy object for the Grappa GlobalStack. The proxy tracks pushes locally in the \texttt{pushed\_values} array. When \texttt{pop} is called, if there are pushes in the local array, it immediately takes the value and wakes the last worker to push. Otherwise, it adds a pointer to a location on its stack where the \texttt{sync} operation, will eventually write the result. Because of local matching, when a Stack proxy is synchronized, it will have either all pops or all pushes, which makes the implementation of sync straightforward. Note that the operation to synchronize a batch of pushes looks almost identical to the code to do a single push from Figure~\ref{fig:push}.

@@

\subsection{Global Set, and Map}
The Grappa GlobalSet uses a simple chaining design, implemented with a global array of cells (allocated from the global heap), which are partitioned evenly among all the cores, and indexed with a hash function. Our flat-combining version supports both \texttt{insert} and \texttt{lookup}. To track all of the keys waiting to be inserted, we use the hash set implementation from the C++11 standard library (\texttt{std::unordered\_set}). Similar to the case with pops, for lookups, we must keep track of pointers for where to put the results of lookup operations, which the proxy does with a hash map (again from the C++ standard library) from keys to a list of pointers. As discussed in Section~\ref{sec:memory-model}, matching lookups with inserts locally would force a particular sequential order, so instead we only allow matching inserts with inserts and lookups with lookups. This allows \texttt{sync} to simply issue all of the inserts and lookups in parallel and block until all have completed.

Our implementation of the GlobalMap matches that of the Set but of course stores values of a given type. This does not change the requirements of the proxy except that the return values of lookups must be a copy of the value rather than just a boolean.

### 2013-06-09 doc/pgas13/grappa.tex
incorporating Mark's feedback and stuff

Grappa is a global-view PGAS-style runtime for commodity clusters which has been designed from the ground up to achieve high performance on irregular applications. The key is latency tolerance---long-latency operations such as remote communication can be tolerated by switching to another concurrent thread of execution. Given abundant  concurrency, there are opportunities to increase throughput by sacrificing latency. In particular, throughput of random accesses to remote memory can be improved by delaying communication requests and aggregating them into larger packets. Highly-tuned implementations of irregular applications in shared-memory, PGAS and message-passing paradigms, typically end up implementing constructs to allow for aggregation and leverage data parallelism. The goal of Grappa is to provide a simple-to-use PGAS abstraction with mechanisms for expressing the concurrency needed to achieve high performance.

The programming interface, implemented as a C++11 library, provides high-level operations to access and synchronize through global shared memory, task spawning capabilities and parallel loop constructs for expressing concurrency. In addition, the Grappa ``standard library'' includes functions to manipulate a global heap as well as provides several synchronized global data structures. The following sections will explain the execution model of Grappa and the current C++ programming interface.

@@

\subsection{Aggregated Communication}
Communication in Grappa is achieved via \emph{active messages}~\cite{vonEicken92} in conjunction with bulk remote direct memory access (RDMA) operations. Various components of Grappa's runtime contribute different kinds of messages to move and modify remote data, perform synchronization, or spawn tasks. To make efficient use of the networks in high-performance systems, which perform better with larger messages (typically 64 KB), all communication in Grappa is sent through an aggregation layer that automatically accumulates messages to the same destination. When enough messages are ready to go to a particular destination node, they are serialized into a buffer and sent using RDMA.

@@

All of these can be addressed by any core in the system using a Grappa \texttt{GlobalAddress}, which encodes both the owning core and the address on that core. Additionally, addresses into the global heap are partitioned in a block-cyclic fashion, so that a large allocation is automatically distributed among many nodes. For irregular applications, this helps avoid hot spots and is typically sufficient for random access.

All accesses to memory owned by another core must be done by that core via a message. In this way, the runtime maintains strict isolation even between processes running on the same physical memory. At the programming level, however, this is hidden behind higher-level remote operations, which in Grappa are called ``delegate'' operations. Figure~\ref{fig:system} shows an example of a delegate \texttt{read()} which blocks the calling task and sends a message to the owner, who sends a reply with the data and wakes the caller.

>>>>

Grappa is a global-view PGAS-style runtime for commodity clusters which has been designed from the ground up to achieve high performance on irregular applications. The key is latency tolerance---long-latency operations such as remote communication can be tolerated by switching to another concurrent thread of execution. Given abundant concurrency, there are opportunities to increase throughput by sacrificing latency. In particular, throughput of random accesses to remote memory can be improved by delaying communication requests and aggregating them into larger packets. Highly-tuned implementations of irregular applications in shared-memory, PGAS and message-passing paradigms, typically end up implementing similar constructs. Grappa includes these as part of its core infrastructure and simply asks the programmer to express concurrency which it can leverage to provide performance.

Grappa's programming interface, implemented as a C++11 library, provides high-level operations to access and synchronize through global shared memory, task spawning capabilities and parallel loop constructs for expressing concurrency. In addition, the Grappa ``standard library'' includes functions to manipulate a global heap as well as provides several synchronized global data structures. The following sections will explain the execution model of Grappa and the current C++ programming interface.

@@

\subsection{Aggregated Communication}
The most basic unit of communication in Grappa is an \emph{active message}~\cite{vonEicken92}.
To make efficient use of the networks in high-performance systems, which typically achieve maximum bandwidth only for messages on the order of 64 KB, all communication in Grappa is sent via an aggregation layer that automatically buffers messages to the same destination.
% Communication in Grappa is achieved via \emph{active messages}~\cite{vonEicken92} in conjunction with bulk remote direct memory access (RDMA) operations. Various components of Grappa's runtime contribute different kinds of messages to move and modify remote data, perform synchronization, or spawn tasks. To make efficient use of the networks in high-performance systems, which perform better with larger messages (typically 64 KB), all communication in Grappa is sent through an aggregation layer that automatically accumulates messages to the same destination. When enough messages are ready to go to a particular destination node, they are serialized into a buffer and sent using RDMA.

@@

All of these can be addressed by any core in the system using a \texttt{GlobalAddress}, which encodes both the owning core and the address on that core. Additionally, addresses into the global heap are partitioned in a block-cyclic fashion, so that a large allocation is automatically distributed among many nodes. For irregular applications, this helps avoid hot spots and is typically sufficient for random access.

All accesses to memory owned by another core must be done by that core via a message, even enforcing strict isolation between processes on the same physical memory. At the programming level, however, this is hidden behind higher-level remote operations, which in Grappa are called ``delegate'' operations. Figure~\ref{fig:system} shows an example of a delegate \texttt{read} which blocks the calling task and sends a message to the owner, who sends a reply with the data and wakes the caller.

### 2013-06-09 doc/pgas13/relatedwork.tex
incorporating Mark's feedback and stuff

Flat combining~\cite{flatCombining} is one of long trail of techniques for allowing a single thread to perform the updates of others. Previous techniques performed combining hierarchically using fixed trees~\cite{yew:combining-trees}, dynamic trees (or ``funnels'')~\cite{funnels}, and randomized trees~\cite{edtrees}. However, the observation of the flat combining work was that in practice, a cleverly-implemented \emph{flat} publication list can outperform more complicated schemes.
Follow-on work introduced parallel flat-combining~\cite{scalableFCQueues}, in which multiple publication lists are combined in parallel and their remainders serialized.
Flat combining was also applied in the NUMA (non-uniform memory access) domain with scalable hierarchical locks~\cite{fcNUMALocks}, which improves on prior work on hierarchical locking by leveraging flat-combining's publication mechanism which amortizes the cost of synchronization.

Our work extends the flat-combining paradigm even further to the PGAS domain, where only software provides the illusion of a physically-distributed memory. The relatively higher cost of communication in PGAS makes flat combining even more compelling. In addition, the physical isolation per node combined with Grappa's engineered per-core isolation guarantees introduce a novel opportunity to apply combining, which led to our ``combining proxy'' design.

In the distributed memory and PGAS domains, implementing scalable and distributed synchronization is the name of the game.
Besides Grappa, other work has proposed leveraging latency tolerance in PGAS runtimes, including MADNESS~\cite{shet:async-upc} which proposes ways of using asynchronous operations in UPC, and Zhang et al.~\cite{zhang:barnes-hut} who use asynchronous messages in a UPC Barnes Hut implementation to overlap computation with communication and allow for buffering.
Jose et al. introduced UPC Queues~\cite{jose:upc-queues}, a library which provides explicit queues as a replacement for locks to synchronize more efficiently and allow for buffering communication. This work demonstrates a similar use case for synchronized queues as in our work, and in fact uses them in the Graph500 benchmark as well. In contrast, Grappa performs much of the same message aggregation invisibly to the programmer and even data structure designer, and the flat combining framework described in this work can be used to implement many different data structures for use in synchronizing.

>>>>

Flat combining~\cite{flatCombining} is the latest of many techniques for allowing a single thread to perform the updates of others. Previous techniques performed combining hierarchically using fixed trees~\cite{yew:combining-trees}, dynamic trees (or ``funnels'')~\cite{funnels}, and randomized trees~\cite{edtrees}. However, the observation of the flat combining work was that in practice, a cleverly-implemented \emph{flat} publication list can outperform more complicated schemes.
Follow-on work introduced parallel flat-combining~\cite{scalableFCQueues}, in which multiple publication lists are combined in parallel and their remainders serialized.
Flat combining was also applied in the NUMA (non-uniform memory access) domain with scalable hierarchical locks~\cite{fcNUMALocks}, which improves on prior work on hierarchical locking by leveraging flat-combining's publication mechanism which amortizes the cost of synchronization.

Our work extends the flat-combining paradigm even further to the PGAS domain, where only software provides the illusion of a physically-distributed memory. The relatively higher cost of communication in PGAS makes flat combining even more compelling. In addition, the physical isolation per node combined with Grappa's engineered per-core isolation guarantees, introduce a novel opportunity to apply combining, which led to our ``combining proxy'' design.

In the distributed memory and PGAS domains, implementing scalable and distributed synchronization is the name of the game.
Besides Grappa, other work has proposed leveraging latency tolerance in PGAS runtimes, including MADNESS~\cite{shet:async-upc} which proposes ways of using asynchronous operations in UPC, and Zhang et al.~\cite{zhang:barnes-hut} who use asynchronous messages in a UPC Barnes Hut implementation to overlap computation with communication and allow for buffering.
Jose et al. introduced UPC Queues~\cite{jose:upc-queues}, a library which provides explicit queues as a replacement for locks to synchronize more efficiently and allow for buffering communication. This work demonstrates a similar use case for synchronized queues as in our work, and uses them with the Graph500 benchmark as well. In contrast, Grappa performs much of the same message aggregation invisibly to the programmer and even data structure designer, and the flat combining framework described in this work can be used to implement many different data structures for use in synchronizing.
